---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  warning = FALSE,
  message = FALSE
)
```

# funspotr

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

The goal of funspotr is to make it easy to identify which functions and packages are used in files.

`{funspotr}` may not identify *every* function and/or package. See [Limitations, Problems, Musings] or read the source code for details.

## Spot functions in a file

The primary function in `{funspotr}` is `spot_funs()` which returns a dataframe showing the functions and associated packages used in a file.

```{r}
library(funspotr)
library(dplyr)

file_lines <- "
library(dplyr)
require(tidyr)

as_tibble(mpg) %>% 
  mutate(class = as.character(class)) %>%
  group_by(class) %>%
  nest() %>%
  mutate(stats = purrr::map(data,
                            ~lm(cty ~ hwy, data = .x)))
 
made_up_fun()
"

file_output <- tempfile(fileext = ".R")
writeLines(file_lines, file_output)

spot_funs(file_path = file_output)
```

* `funs`: function used in script
* `pkgs`: `{funspotr}`s best guess as to the package it came from
* `in_multiple_pkgs` : Whether the function has multiple packages/environments on it's search space. By default only the package at the top of the search space is  returned^[E.g. `as_tibble()` is attributed to `{tidyr}` by `spot_funs()` however `as_tibble()` is also in `{dplyr}`. I don't worry about getting to the root source of the package or the fact that both of those packages are just reexporting it from `{tibble}`. Setting `keep_search_list = TRUE` will return rows for each item in the search list which may be helpful if getting unexpected results.)]. 

<!-- The example below uses `spot_pkgs_from_DESCRIPTION()` to load in package dependencies and then passes the resulting character vector to `spot_funs_custom()`. -->

`spot_funs_custom()` allows the user to explicitly specify `pkgs` where functions may come from. This is useful in cases where the packages loaded are not in the same location as the `file_path` (e.g. they are loaded via `source()` or a DESCRIPTION file, or some other workflow). For example, below is a made-up example where the `library()` calls are made in a separate file and `source()`d in.

```{r}
# file where packages are loaded
file_libs <- "library(dplyr)
library(lubridate)"

file_libs_output <- tempfile(fileext = ".R")
writeLines(file_libs, file_libs_output)

# File of interest where things happen
file_run <- glue::glue(
"source('{ file_libs_output }')
  
tibble::tibble(days_from_today = 0:10) %>% 
    mutate(date = today() + days(days_from_today))
", 
file_libs_output = stringr::str_replace_all(file_libs_output, "\\\\", "/")
)

file_run_output <- tempfile(fileext = ".R")
writeLines(file_run, file_run_output)

# Identify packages using both files and then pass in explicitly to `spot_funs_custom()`
pkgs <- c(spot_pkgs(file_libs_output), 
          spot_pkgs(file_run_output, show_explicit_funs = TRUE))

spot_funs_custom(
  pkgs = pkgs,
  file_path = file_run_output)
```

## Spot functions in a github repo

`github_spot_funs()` is a wrapper on `spot_funs()` that maps it across all R or Rmarkdown files in a github repository.

```{r}
# repo for an old presentation I gave
gh_ex <- github_spot_funs(
  repo = "brshallo/feat-eng-lags-presentation", 
  branch = "main")

gh_ex
```

* `contents` : filepath in github repo
* `urls`: URL to raw file on github
* `spotted`: List-column of results when mapping `spot_funs()` across `urls`^[If any file could not be parsed will be  a `purrr::safely()` list-column output where each item is a list containing `result` and `error`.]. 

These results may then be unnested with the helper `funspotr::unnest_github_results()` to provide a table of which functions and packages are used. You can manipulate this -- for example say we want to filter to those places where `{here}`, `{readr}` or `{rsample}` packages are used.

```{r}
gh_ex %>% 
  unnest_github_results() %>% 
  filter(pkgs %in% c("here", "readr", "rsample"))
```

You can set `preview = TRUE` in `github_spot_funs()` to first view the files you intend to parse.

```{r}
preview_files <- github_spot_funs(
  repo = "brshallo/feat-eng-lags-presentation", 
  branch = "main",
  preview = TRUE)

preview_files
```

You can pass in a custom set of urls by inputting a dataframe to the argument `custom_urls`. E.g. say we only want to parse the "types-of-splits.R" and "Rmd-to-R.R" files.

```{r}
files <- preview_files %>% 
  filter(stringr::str_detect(contents, "types-of-splits|Rmd-to-R"))

github_spot_funs(custom_urls = files) %>% 
  unnest_github_results()
```

## Files you didn't write

Functions created in a file and functions from unavailable packages or packages that don't exist will output as `pkgs = "(unknown)"`. 

```{r}
file_lines_missing_pkgs <- "
library(dplyr)

as_tibble(mpg)

hello_world <- function() print('hello world')

madeuppkg::made_up_fun()

hello_world()
"

missing_pkgs_ex <- tempfile(fileext = ".R")
writeLines(file_lines_missing_pkgs, missing_pkgs_ex)

spot_funs(file_path = missing_pkgs_ex)
```

*To spot which package a function is from you must have the package installed locally.* Hence for files on other peoples' github repositories or that you created on a different machine, it is a good idea to start with `funspotr::check_pkgs_availability()` to see which packages you are missing.

A quick helper for installing missing packages (see "R/spot-pkgs.R" for documentation):

```r
check_pkgs_availability(file_path) %>% 
  funspotr:::install_missing_pkgs()
```

Alternatively, you may want to clone the repository locally and then use `renv::dependencies()` to identify and install the package dependencies to your local repository and only then start using `{funspotr}`. ([renv](https://rstudio.github.io/renv/) is a more robust approach to finding and installing dependencies than using the `{funspotr}` functions -- particularly in cases where you are missing many dependencies or don't want to alter the packages in your global library.)

# Other Things

## Show all function calls

Passing in `show_each_use = TRUE` to `...` in `get_funs()` or `github_spot_funs()` will return *all* instances of a function call rather than just once for each file. 

Note `mutate()` at both rows 4 and 8:
```{r}
spot_funs(file_path = file_output, show_each_use = TRUE)
```

## Helper for `{blogdown}` tags

Setting `as_yaml_tags = TRUE` in `spot_pkgs() flattens the dependencies and outputs them in a format that can be pasted into the **tags** section of a blogdown post's YAML header. See ([blogdown#647](https://github.com/rstudio/blogdown/issues/647)).

```{r}
# Example from old blogdown post
spot_pkgs(
  file_path = "https://raw.githubusercontent.com/brshallo/brshallo/master/content/post/2020-02-06-maximizing-magnetic-volume-the-perfect-bowl.Rmd",
  as_yaml_tags = TRUE)
```

## Unexported functions

Many of the unexported functions (in "R/github-spot.R" in particular) may be helpful in building up other workflows for mapping `spot_funs()` across multiple files. (Most unexported functions in `funspotr` still include documentation). *If you have a suggestion for a function or set of functions for spotting functions, feel free to open an issue.* 

<!-- **If you've used {funspotr} to map the R functions and packages of a public blog or repository, open an issue to add a link in the README.** -->

# How `funspotr::spot_funs()` works

At a high-level...

1. Create a new R instance using [callr](https://github.com/r-lib/callr)
2. Load packages. Explicit calls (e.g. `pkg::fun()`) are loaded individually via [import](https://github.com/rticulate/import) and are loaded last (putting them at the top of the search space)^[This heuristic is imperfect and means that a file with "library(dplyr); select(); MASS::select()" would view both `select()` calls as coming from `{MASS}` -- when what it should do is view the first was as coming from {dplyr} and the second from {MASS}.].

(steps 1 and 2 needed so that later `find()` step will have best chance of identifying the package a function comes from in the file.)

3. Pass file through `utils::getParseData()` and filter to just functions
4. Pass functions through `utils::find()` to identify associated package

# Limitations, Problems, Musings

* If a file contains R syntax that is not well defined it will not be parsed
* `knitr::read_chunk()` and `knitr::purl()` will also often frequently break parsing. See [knitr#1753](https://github.com/yihui/knitr/issues/1753) & [knitr#1938](https://github.com/yihui/knitr/issues/1938)
* Please open an issue if you find other cases where it breaks :-) .
* As mentioned elsewhere, the default parsing of `spot_funs()` and functions that build on this is primarily for cases where package dependencies are loaded in the same file that they are used in^[i.e. in interactive R scripts or Rmd documents where you use `library()` or related calls within the script.]. Scripts that are not self-contained typically should have the `pkgs` argument provided explicitly via `spot_funs_custom()`.
* Feel free to open an issue if you'd be interested in a simplifying function or vignette for mapping `spot_funs()` through local repos or other folder structures other than github respositories (as in `github_spot_funs()`)^[E.g. for mapping across a local repository, across a package, a data workflow, etc.]
* All the functions in "R/spot-pkgs.R" would probably be better handled by something like `renv::dependencies()` or a parsing based approach. The simple regex's I use have a variety of problems. As just one example `funspotr::get_pkgs()` will not recognize when a package is within quotes or being escaped^[e.g. in this case `lines <- "library(pkg)"`  the `pkg` would show-up as a dependency despite just being part of a quote rather than actually loaded.].
* I am curiuos if there is something to be learned from how `R CMD check` does function parsing. 
    * `{funspotr}`s urrent approach is slow
    * Current approach uses some imperfect heuristics
* Does not identify infix operators, e.g. `+` (maybe this is fine though)
* `funspotr` has LOTS of dependencies
* It may have made more sense to have `github_get_funs()` clone the repo locally rather than pointing to URL's to parse each file.
* Currently is possible to have github block you pretty soon due to hitting too many files (in which case you'll likely get a 403 or connection error). There are some things that could probably be done to reduce number of github API hits (e.g. above bullet).
* Throughout the code and package documentation I have "inspiration" bullets followed by a link pointing to places where I took stuff from stack overflow, github, or other packages

## Installation

You can install the development version of funspotr from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("brshallo/funspotr")
```
