---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  warning = FALSE,
  message = FALSE
)
```

# funspotr

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

The goal of funspotr (R function spotter) is to make it easy to identify which functions and packages are used in files. It was initially written to map out the functions and packages used in a few popular github repositories (see [brshallo/funspotr-examples](https://github.com/brshallo/funspotr-examples) and associated [blog post](https://www.bryanshalloway.com/2022/01/18/identifying-r-functions-packages-used-in-github-repos/) and [post on parsing gists](https://www.bryanshalloway.com/2022/02/07/identifying-r-functions-packages-in-your-github-gists/)).

funspotr is primarily designed for self-contained scripts (see [Package dependencies in another file]). Also, it may not identify *every* function and/or package (see [Limitations, Problems, Musings]) or read the source code for details.

## Spot functions in a file

The primary function in funspotr is `spot_funs()` which returns a dataframe showing the functions and associated packages used in a file.

```{r}
library(funspotr)
library(dplyr)

file_lines <- "
library(dplyr)
require(tidyr)

as_tibble(mpg) %>% 
  mutate(class = as.character(class)) %>%
  group_by(class) %>%
  nest() %>%
  mutate(stats = purrr::map(data,
                            ~lm(cty ~ hwy, data = .x)))
 
made_up_fun()
"

file_output <- tempfile(fileext = ".R")
writeLines(file_lines, file_output)

spot_funs(file_path = file_output)
```

* `funs`: function used in script
* `pkgs`: best guess as to the package it came from
* `in_multiple_pkgs` : Whether the function has multiple packages/environments on it's (guessed) search space. By default only the package at the top of the search space is returned^[E.g. `as_tibble()` is attributed to [tidyr](https://tidyr.tidyverse.org/) by `spot_funs()` however `as_tibble()` is also in [dplyr](https://dplyr.tidyverse.org/). I don't worry about getting to the root source of the package or the fact that both of those packages are just reexporting it from [tibble](https://tibble.tidyverse.org/). Setting `keep_search_list = TRUE` will return rows for each item in the search list which may be helpful if getting unexpected results.)]. 

<!-- The example below uses `spot_pkgs_from_DESCRIPTION()` to load in package dependencies and then passes the resulting character vector to `spot_funs_custom()`. -->

## Spot functions in a github repo

`github_spot_funs()` is a wrapper on `spot_funs()` that maps it across all R or Rmarkdown files in a github repository.

```{r}
# repo for an old presentation I gave
gh_ex <- github_spot_funs(
  repo = "brshallo/feat-eng-lags-presentation", 
  branch = "main")

gh_ex
```

* `contents` : filepath in github repo
* `urls`: URL to raw file on github
* `spotted`: `purrr::safely()` style list-column of results when mapping `spot_funs()` across `urls`^[If any file could not be parsed will be  a `purrr::safely()` list-column output where each item is a list containing `result` and `error`.]. 

These results may then be unnested with the helper `funspotr::unnest_github_results()` to provide a table of which functions and packages are used. This can be manipulated like any other dataframe -- say we want to filter to those files where [here](https://here.r-lib.org/), [readr](https://readr.tidyverse.org/) or [rsample](https://rsample.tidymodels.org/) packages are used.

```{r}
gh_ex %>% 
  unnest_github_results() %>% 
  filter(pkgs %in% c("here", "readr", "rsample"))
```

The outputs from `funspotr::unnest_github_results()` can also be passed into `funspotr::network_plot()` to visualize the connections between functions/packages and files.

### Previewing and customizing files to parse

You can set `preview = TRUE` in `github_spot_funs()` to first view the files you intend to parse.

```{r}
preview_files <- github_spot_funs(
  repo = "brshallo/feat-eng-lags-presentation", 
  branch = "main",
  preview = TRUE)

preview_files
```

You can pass in a custom set of urls by inputting a dataframe to the argument `custom_urls` -- this may be the output of a previous call where `preview = TRUE`. For examle, say we only want to parse the "types-of-splits.R" and "Rmd-to-R.R" files.

```{r}
preview_files %>% 
  filter(stringr::str_detect(contents, "types-of-splits|Rmd-to-R")) %>% 
  github_spot_funs(custom_urls = .) %>% 
  unnest_github_results()
```

The `custom_urls` argument works for any situation where you want to map `spot_funs()` across a set of filepaths or URLs. E.g. see example in `github_gists()` for a case where this argument is used to parse functions/packages from a specified github user's gists.

## Files you didn't write

Functions created in the file as well as functions from unavailable packages (or packages that don't exist) will output as `pkgs = "(unknown)"`. 

```{r}
file_lines_missing_pkgs <- "
library(dplyr)

as_tibble(mpg)

hello_world <- function() print('hello world')

madeuppkg::made_up_fun()

hello_world()
"

missing_pkgs_ex <- tempfile(fileext = ".R")
writeLines(file_lines_missing_pkgs, missing_pkgs_ex)

spot_funs(file_path = missing_pkgs_ex)
```

*To spot which package a function is from you must have the package installed locally.* Hence for files on others' github repos or that you created on a different machine, it is a good idea to start with `funspotr::check_pkgs_availability()` to see which packages you are missing.

A quick helper for installing missing packages (see "R/spot-pkgs.R" for documentation):

```r
check_pkgs_availability(file_path) %>% 
  funspotr:::install_missing_pkgs()
```

Alternatively, you may want to clone the repository locally and then use `renv::dependencies()` and only then start using funspotr. [renv](https://rstudio.github.io/renv/) is a more robust approach to finding and installing dependencies -- particularly in cases where you are missing many dependencies or don't want to alter the packages in your global library.

# Other Things

## Package dependencies in another file

`spot_funs()` is currently set-up for self-contained files. But `spot_funs_custom()` allows the user to explicitly specify `pkgs` where functions may come from. This is useful in cases where the packages loaded are not in the same location as the `file_path` (e.g. they are loaded via `source()` or a DESCRIPTION file, or some other workflow). For example, below is a made-up example where the `library()` calls are made in a separate file and `source()`d in.

```{r}
# file where packages are loaded
file_libs <- "library(dplyr)
library(lubridate)"

file_libs_output <- tempfile(fileext = ".R")
writeLines(file_libs, file_libs_output)

# File of interest where things happen
file_run <- glue::glue(
"source('{ file_libs_output }')
  
tibble::tibble(days_from_today = 0:10) %>% 
    mutate(date = today() + days(days_from_today))
", 
file_libs_output = stringr::str_replace_all(file_libs_output, "\\\\", "/")
)

file_run_output <- tempfile(fileext = ".R")
writeLines(file_run, file_run_output)

# Identify packages using both files and then pass in explicitly to `spot_funs_custom()`
pkgs <- c(spot_pkgs(file_libs_output), 
          spot_pkgs(file_run_output, show_explicit_funs = TRUE))

spot_funs_custom(
  pkgs = pkgs,
  file_path = file_run_output)
```

Also see `funspotr:::spot_pkgs_from_DESCRIPTION()`.

## Show all function calls

Passing in `show_each_use = TRUE` to `...` in `get_funs()` or `github_spot_funs()` will return *all* instances of a function call rather than just once for each file. 

Compared to the initial example, `mutate()` now shows-up at both rows 4 and 8:
```{r}
spot_funs(file_path = file_output, show_each_use = TRUE)
```

## Helper for [blogdown](https://pkgs.rstudio.com/blogdown/) tags

Setting `as_yaml_tags = TRUE` in `spot_pkgs()` flattens the dependencies and outputs them in a format that can be copied and pasted into the **tags** section of a blogdown post's YAML header.

```{r}
# Example from old blogdown post
spot_pkgs(
  file_path = "https://raw.githubusercontent.com/brshallo/brshallo/master/content/post/2020-02-06-maximizing-magnetic-volume-the-perfect-bowl.Rmd",
  as_yaml_tags = TRUE) %>% 
  cat()
```

`spot_pkgs_used()` will only return those packages that have functions actually used^[E.g. for cases when there are library calls that aren't actually used in the file. This may be useful in cases when metapackages like tidyverse or tidymodels are loaded but not all packages are actually used.].

To automatically have your packages used as the tags for a post you can add the function `funspotr::spot_tags()` to the `tags` argument of your YAML header. For example:

```
---
title: This is a post
author: brshallo
date: '2022-02-11'
tags: ["`r funspotr::spot_tags(drop_knitr = TRUE)`"]
slug: this-is-a-post
---
```

See ([blogdown#647](https://github.com/rstudio/blogdown/issues/647#issuecomment-1041599327)) for an explanation.

## Unexported functions

Many of the unexported functions (in "R/github-spot.R" in particular) may be helpful in building up other workflows for mapping `spot_funs()` across multiple files^[Most unexported functions in `funspotr` still include a man file and at least partial documentation.] *If you have a suggestion for a function or set of functions for spotting functions, feel free to open an issue.* 

<!-- **If you've used {funspotr} to map the R functions and packages of a public blog or repository, open an issue to add a link in the README.** -->

# How `funspotr::spot_funs()` works

At a high-level...

1. Create a new R instance using [callr](https://github.com/r-lib/callr)
2. Load packages. Explicit calls (e.g. `pkg::fun()`) are loaded individually via [import](https://github.com/rticulate/import) and are loaded last (putting them at the top of the search space)^[This heuristic is imperfect and means that a file with "library(dplyr); select(); MASS::select()" would view both `select()` calls as coming from {MASS} -- when what it should do is view the first was as coming from {dplyr} and the second from {MASS}.].

(steps 1 and 2 needed so that step 4 has the best chance of identifying the package a function comes from in the file.)

3. Pass file through `utils::getParseData()` and filter to just functions (inspired by `NCmisc::list.functions.in.file()`)
4. Pass functions through `utils::find()` to identify associated package

# Limitations, Problems, Musings

* If a file contains R syntax that is not well defined it will not be parsed and will return an error. See [formatR](https://yihui.org/formatr/#6-further-notes) (used by {funspotr} in parsing) for other common reasons for failure.
* `knitr::read_chunk()` and `knitr::purl()` in a file passed to {funspotr} will also frequently cause an error in parsing. See [knitr#1753](https://github.com/yihui/knitr/issues/1753) & [knitr#1938](https://github.com/yihui/knitr/issues/1938)
* Please open an issue if you find other cases where it breaks :-) .
* As mentioned elsewhere, the default parsing of `spot_funs()` is primarily for cases where package dependencies are loaded in the same file that they are used in^[i.e. in interactive R scripts or Rmd documents where you use `library()` or related calls within the script.]. Scripts that are not self-contained typically should have the `pkgs` argument provided explicitly via `spot_funs_custom()`.
* funspotr does not pay attention to when functions are reexported from elsewhere. For example, many tibble functions are reexported by dplyr and tidyr -- funspotr though will not know the true home of these functions it is simply looking at the top of the search space^[For example when reviewing David Robinson's Tidy Tuesday code I found that the [meme](https://github.com/GuangchuangYu/meme) package was used far more than I would have expected. Turns out it was just due to it reexporting the aes() function from ggplot.].
* Feel free to open an issue if you'd be interested in a simplifying function or vignette for mapping `spot_funs()` through local repos or other folder structures other than github respositories (which is already covered in `github_spot_funs()`)^[E.g. for mapping across a local repository, across a package, a data workflow, etc.]
* All the functions in "R/spot-pkgs.R" would probably be better handled by something like `renv::dependencies()` or a parsing based approach. The simple regex's I use have a variety of problems. As just one example `funspotr::get_pkgs()` will not recognize when a package is within quotes or being escaped^[e.g. in this case `lines <- "library(pkg)"`  the `pkg` would show-up as a dependency despite just being part of a quote rather than actually loaded.].
* I am curiuos if there is something to be learned from how `R CMD check` does function parsing. 
    * `funspotr's current approach is slow
    * Current approach uses some imperfect heuristics
* Does not identify infix operators, e.g. `+` (maybe fine though)
* `funspotr` has lots of dependencies. It may have make sense to move some of the non-core functionality into a separate package (e.g. everything "R/github-spot.R")
* It may have made more sense to have `github_get_funs()` clone the repo locally rather than pointing to URL's to parse each file.
* Currently is possible to have github block you pretty soon due to hitting too many files (in which case you'll likely get a 403 or connection error). There are some things that could probably be done to reduce number of github API hits (e.g. above bullet, `Sys.sleep()`, ...).
* Throughout the code and package documentation I have "inspiration" bullets followed by a link pointing to places where I took stuff from stack overflow, github, or other packages

## Installation

You can install the development version of funspotr from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("brshallo/funspotr")
```
