---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# funspotr

<!-- badges: start -->
<!-- badges: end -->

The goal of funspotr is to make it easy to identify which functions and associated packages are used in files.

`{funspotr}` may not identify *every* function and/or package used in scripts. See [Limitations, Problems, Musings] (or read the source code).

## Installation

You can install the development version of funspotr from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("brshallo/funspotr")
```

## Spot functions in a file

The foundational function in `{funspotr}` is `spot_funs()` which returns a dataframe showing the functions and associated packages used in a single file.

```{r}

```

`spot_funs_custom()` allows the user to explicitly specify `pkgs` where functions may come from. This is useful in cases where the packages loaded are not in the same location as the `file_path` (e.g. they are loaded via `source()` or a DESCRIPTION file). 

The example below uses `spot_pkgs_from_DESCRIPTION()` to load in package dependencies and then passes the resulting character vector to `spot_funs_custom()`.

```{r}
# use funspotr as example

```

## Spot functions in a github repo

`github_spot_funs()` identifies the R and Rmarkdown files in a github repository and then maps `spot_funs()` across all file URLs, outputting the results to a `spotted` list-column in a dataframe.

```{r}

```

These results may then be unnested to provide a table of which functions and packages are used in each file.

```{r}

```

You can set `preview = TRUE` in `github_spot_funs()` to first view the files you intend to parse and spot functions in. (This is particularly helpful in repositories with many files where it will take a while for `github_spot_funs()` to run.)

```{r}

```

## Files you didn't write

Functions created in a file and functions from unavailable packages will output as `pkgs = "(unknown)"`. *To spot which package a function is from you must have the package installed locally.*

Hence for files on other peoples' github repositories or that you created on a different machine, it is a good idea to start with `funspotr::check_pkgs_availability()` to see which packages you may be missing. 

`check_pkgs_availability(file_path) %>% funspotr:::install_missing_pkgs()` is a quick helper for installing missing packages (see "R/spot-pkgs.R" for documentation). 

Alternatively, you may want to clone the repository locally and then use `renv::dependencies()` to identify and install the package dependencies to your local repository (see [{renv}](https://rstudio.github.io/renv/) for documentation) and only then start using `{funspotr}`. 

(`{renv}` is a more robust approach than using the `{funspotr}` functions and is particular important in cases where you are missing many dependencies or don't want to alter the packages in your global library).

# Other Things

**Helper for `{blogdown}` tags:**

`spot_pkgs()` identifies packages loaded in an individual file and has an argument for `as_yaml_tags` that flattens the dependencies and outputs them in a format that can be pasted into the **tags** section of a blogdown post's YAML header. See ([blogdown#]())

```{r}
```

**Unexported functions:**

Many of the unexported functions (in "R/github-spot.R" in particular) may be helpful in building up other workflows for mapping `spot_funs()` across multiple files. (Most unexported functions in `funspotr` still include documentation).

*If you have a suggestion for a function or set of functions for spotting functions, feel free to open an issue.* 

**If you've used {funspotr} to map the R functions and packages of a public blog or repository, open an issue to add a link in the README.**


## Limitations, Problems, Musings

* As mentioned elsewhere, the current parsing is primarily for cases where the packages are loaded in the same file that they are used in (i.e. in interactive scripts or notebooks). Scripts that are not self-contained typically require passing in a `pkgs` argument explicitly to `spot_funs_custom()`.
* Feel free to open an issue if you'd be interested in a simplifying function or vignette for mapping `spot_funs()` through local repos or other folder structures outside of github repos containing largely self-contained files as in `github_spot_funs()`. E.g. for mapping across a local repository, across a package, a data workflow, etc.
* All the functions in "R/spot-pkgs.R" would be better handled by something like `renv::dependencies()` or a parsing based approach. The simple regex's I use have a variety of problems. As just one example `funspotr::get_pkgs()` will not recognize when a package is within quotes or being escaped, e.g. in this case `lines <- "library(pkg)"`  the `pkg` would show-up as a dependency despite just being part of a quote rather than actually loading `pkg`.
* The foundation of "R/spot-funs.R" is `NCmisc::list.functions.in.file()` which would likely be better if instead handled by however `R CMD check` does function parsing. 
    * The current approach is much slower than `R CMD check`
    * `NCmisc::list.functions.in.file()` does not properly consider namespaces -- making the `in_multiple_pkgs` column in the output of `spot_funs()` necessary in cases of ambigugity (a better handling of namespaces should obviate the need for this column).
        * Even when there is an explicit call to a function (e.g. `pkg::fun()`), `NCmisc::list.functions.in.file()` will not know `fun` is coming from `pkg` specifically (if there is another package loaded elsewhere that also contains a function named `fun()`)
    * Does not identify infix operators, e.g. `+` (maybe this is fine though)
* `funspotr` has LOTS of dependencies
* It may have made more sense to have `github_get_funs()` clone the repo locally rather than pointing to URL's for each parsed file.

